Wasted Money? Changes in Toronto's Annual Police Budget and Reported

### 2024-09-27 paper/paper.qmd
Added content and citations

Using official data from Open Data Toronto @opendatatoronto and the R programming language @r, the annual police budget, the number of reported crimes, and the number of cleared crimes were analyzed for the years 2020 to 2023. The data revealed that the number of reported crimes consistently rose while the number of cleared crimes remained relatively stable, leading to a reduction in the clearance ratio. The police department's total annual budget increased slightly from 2020 to 2022 but fell by over 45% in 2023. The drastic change in the police budget and the consistent crime numbers indicate that the Toronto police have little influence on crime rates. Therefore, this paper encourages a thorough analysis of why investment in Toronto's police department has not resulted in fewer crimes and increased safety.

@@

# Data {#sec-mydatasection}
## Introduction of Datasets
Two datasets were extracted from Open Data Toronto @opendatatoronto: one records all instances of expenditure and revenue in Toronto Police since 2020, and the other contains the number of reported crimes in various categories along with the corresponding number of cleared crimes up to the end of 2023. Summary statistics of the disaggregated data are presented in @tbl-raw-bud and @tbl-raw-cri in the Appendix. These tables are excluded from the main body because the analysis focuses on aggregate budgets and total crime counts rather than individual expenses and specific crime categories.

Open Data Toronto was chosen as the data source because it provides official data from the City of Toronto, ensuring credibility, variety, and depth. The only alternative source for crime and police budget data is the Toronto Police data portal, and its datasets are identical to those in Open Data Toronto.

The data measurements from @opendatatoronto are not entirely reliable. The police budget dataset includes approved budgets from the Toronto Police Services Board and City Council, as well as operational expenses incurred. Expenses or revenues that were not officially reported are excluded from the dataset. The crime reports and cleared crimes dataset relies on reports filed by officers, meaning undiscovered crimes are not included, and unfounded reports are counted. These measurement biases complicate trend interpretation. For example, an increase in reported crimes may reflect either an actual rise in crime or an improvement in officers' ability to detect crimes.

The cleaned, aggregated datasets generated using knitr @knitr and KableExtra @kextra are as follows:

@@

Data cleaning and calculations were performed to derive the variables in @tbl-bud and @tbl-cri. The "Year" variable was limited to 2020–2023 because the police budget dataset begins in 2020 and the crime report dataset extends to 2023. "Yearly Budget" was calculated by summing all individual expenses and revenues in the disaggregated dataset, with expenses represented as positive values and revenues as negative values. "Change From Last" was determined by subtracting the previous year's "Yearly Budget" from the current year, excluding 2020, which has no preceding year. "Percent Change" was calculated by dividing the previous year's "Yearly Budget" by the current "Change From Last." For @tbl-cri, "Yearly Crimes" and "Cleared Crimes" were calculated by summing the reported and cleared cases across all crime categories, respectively. The data cleaning process was facilitated by the R package tidyverse @tidyverse.

@@

>>>>

Using official data from Open Data Toronto [@opendatatoronto] and the R programming language [@r], the annual police budget, the number of reported crimes, and the number of cleared crimes were analyzed for the years 2020 to 2023. The data revealed that the number of reported crimes consistently rose while the number of cleared crimes remained relatively stable, leading to a reduction in the clearance ratio. The police department's total annual budget increased slightly from 2020 to 2022 but fell by over 45% in 2023. The drastic change in the police budget and the consistent crime numbers indicate that the Toronto police have little influence on crime rates. Therefore, this paper encourages a thorough analysis of why investment in Toronto's police department has not resulted in fewer crimes and increased safety.

@@

# Data {#sec-mydatasection}
## Introduction of Datasets
Two datasets were extracted from Open Data Toronto [@opendatatoronto]: "Toronto Police Budget" records all instances of expenditure and revenue in Toronto Police since 2020, and "Police Annual Statistical Report - Reported Crimes" contains the number of reported crimes in various categories along with the corresponding number of cleared crimes up to the end of 2023. Summary statistics of the disaggregated data are presented in @tbl-raw-bud and @tbl-raw-cri in the Appendix. These tables are excluded from the main body because the analysis focuses on aggregate budgets and total crime counts rather than individual expenses and specific crime categories.

Open Data Toronto was chosen as the data source because it provides official data from the City of Toronto, ensuring credibility, variety, and depth. The city-wide datasets used for this study were difficult to gather, as the aggregation of individual cases requires effort from all members of Toronto Police. Data on reported crimes and clearance rates are also essential measures of public safety and police performance. Thus, to ensure the study's authenticity, official datasets are considered the only reliable sources. The only alternative source for crime and police budget data is the Toronto Police data portal, and its datasets are identical to those in Open Data Toronto.

The data measurements from @opendatatoronto are not entirely reliable. The police budget dataset includes approved budgets from the Toronto Police Services Board and City Council, as well as operational expenses incurred. Expenses or revenues that were not officially reported are excluded from the dataset. The crime reports and cleared crimes dataset relies on reports filed by officers, meaning undiscovered crimes are not included, and unfounded reports are counted. These measurement biases complicate trend interpretation. For example, an increase in reported crimes may reflect either an actual rise in crime or an improvement in officers' ability to detect crimes. Another apparent limitation of the datasets is the time frame; the two datasets only overlap between 2020 and 2023, leaving minimal room for temporal analysis. Time related variables in the raw datasets do not store months or dates, preventing detailed temporal comparison.

The cleaned, aggregated datasets generated using 'knitr' [@knitr] and 'KableExtra' [@kextra] are as follows:

@@

Data cleaning and calculations were performed to derive the variables in @tbl-bud and @tbl-cri. The "Year" variable was limited to 2020–2023 because the police budget dataset begins in 2020 and the crime report dataset extends to 2023. "Yearly Budget" was calculated by summing all individual expenses and revenues in the disaggregated dataset, with expenses represented as positive values and revenues as negative values. "Change From Last" was determined by subtracting the previous year's "Yearly Budget" from the current year, excluding 2020, which has no preceding year. "Percent Change" was calculated by dividing the previous year's "Yearly Budget" by the current "Change From Last." For @tbl-cri, "Yearly Crimes" and "Cleared Crimes" were calculated by summing the reported and cleared cases across all crime categories, respectively. The data cleaning process was facilitated by the R package 'tidyverse' [@tidyverse]. All scripts were formatted using the package 'styler' [@styler].

@@

```
\vspace{1em}
\FloatBarrier
